---
title: 'ICU Data Analysis'
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---
```{r}
#load data 
load("icu.Rdata")

#load libraries
library(tidyverse)
```
```{r}
#view the data
glimpse(icu)
```

Preliminary Visualizations of Variables  
```{r}
#Independent Variable
table(icu$vital.status)

#Demographics 
table(icu$age)
table(icu$race)
table(icu$gender)

#Plots of demographics and y-variable
barplot(table(icu$vital.status))
barplot(table(icu$age))
barplot(table(icu$race))
barplot(table(icu$gender))
```
The study population is primarily white males between the ages of 55 and 75. The majority of patients also survived after their visit to the ICU. 

The percentages of patients in each demographic category have been calculated below:
```{r}
#Race : White, Black, or other
race_stats <- c(175/200, 15/200, 10/200)
race_stats

#Gender : Male or Female
gender_stats <- c(124/200, 76/200)
gender_stats

#Vital Status : Lived or Died
vital_stats <- c(160/200, 40/200)
vital_stats
```
It can be concluded that 87.5% of the patients are white whereas only 7.5% are black. The calculations additionally show us that 62% are male and 38% of the patients are female. Lastly, 80% of the ICU patients survived following their hospital discharge and 20% died. 


```{r}
#General factors
table(icu$conscious)
table(icu$previous)
table(icu$service)
table(icu$type)

#Barplots of general factors
barplot(table(icu$conscious))
barplot(table(icu$previous))
barplot(table(icu$service))
barplot(table(icu$type))
```
The percentages of patients in each general category have been calculated below:

```{r}
#Conscious : No coma/stupor, deep stupor, or coma
cons_stats <- c(185/200, 5/200, 10/200)
cons_stats

#Previous : No or yes
prev_stats <- c(170/200, 30/200)
prev_stats

#Service : Medical or Surgical
serv_stats <- c(93/200, 107/200)
serv_stats

#Type : Elective or Emergency
type_stats <- c(53/200, 147/200)
type_stats
```
Based off of the calculations it can be concluded that 92.5% of patients didn't have a coma and 2.5% had a deep stupor when they entered the ICU. Additionally, 85% of patients visited the ICU in the past 6 months and 15% did not. Lastly, 46.5% of patients had a medical service, 53.5% had a surgical service, 26.5% had an elective admission, and 73.5% had an emergency admission. 

```{r}
#Arterial Blood Gas factors
table(icu$PH)
table(icu$PCO2)
table(icu$bicarb)
table(icu$PO2)

#Plots of medical factors
barplot(table(icu$PH))
barplot(table(icu$PCO2))
barplot(table(icu$bicarb))
barplot(table(icu$PO2))
```
Based off of the visualizations it can be concluded that the majority of patients did not have cancer or a fracture. The majority also were within the healthy levels of pH, PCO2, PO2, and bicarbonate.  

Study the relationships between the variables:
```{r}
#Plot the relationship between vital status and age
plot(icu$vital.status, icu$age)
```
It is important to note that the vital.status variable is binary and the distribution of the age of people in this data set is highest between the ages of 55-75. This is important when analyzing the vital status variable because there is a largely different frequency for each age group. Based on the plot between both variables, it is clear that as people get older, their chance of dying is much higher when compared to patients that are younger. We can see that the mean of people that died is roughly 70 whereas, the mean the people that lived is closer to 60. It is also important to note that there are two outliers in the people that died. This again shows that it was rare to have very young people die.    

```{r}
#Plot the relationship between vital status and gender
icu %>% 
  select(vital.status, gender) %>% 
  ggplot() + 
  geom_bar(aes(x=vital.status)) +
  facet_grid(~gender) +
  labs(title = "Vital Status Depending on Gender")
```
This graph illustrates a patient's vital status depending on their gender. It can be concluded that more males survived compared to females. 
```{r}
#Plot the relationship between vital status and race
icu %>% 
  select(vital.status, race) %>% 
  ggplot() + 
  geom_bar(aes(x=vital.status)) +
  facet_grid(~race) +
  labs(title = "Vital Status Depending on Race")
```
It can be concluded that more white people survived after visiting the ICU when compared to other races. It is also important to recognize that there is a large difference between the number of white patients compared to minority groups. This large numerical difference in distributions can largely impact the outcome. 

```{r}
#Plot the relationship between vital status and pH
icu %>% 
  select(vital.status, PH) %>% 
  ggplot() + 
  geom_bar(aes(x=vital.status)) +
  facet_grid(~PH) +
  labs(title = "Vital Status Depending on pH")
```
This graph shows us that patients who had a pH value greater than 7.25 had a higher chance of survival. 
```{r}
#Plot the relationship between vital status and PC02
icu %>% 
  select(vital.status, PCO2) %>% 
  ggplot() + 
  geom_bar(aes(x=vital.status)) +
  facet_grid(~PCO2) +
  labs(title = "Vital Status Depending on PCO2")
```
This graph shows us that patients who had a PCO2 value greater than 45 had a higher chance of survival.

```{r}
#Plot the relationship between vital status and bicarb
icu %>% 
  select(vital.status, bicarb) %>% 
  ggplot() + 
  geom_bar(aes(x=vital.status)) +
  facet_grid(~bicarb) +
  labs(title = "Vital Status Depending on Bicarbonate Levels")
```
This graph shows us that patients who had a bicarbonate levels greater than 18 had a higher chance of survival.
```{r}
#Plot the relationship between vital status and PO2
icu %>% 
  select(vital.status, PO2) %>% 
  ggplot() + 
  geom_bar(aes(x=vital.status)) +
  facet_grid(~PO2) +
  labs(title = "Vital Status Depending on PO2")
```
This graph shows us that patients who had a PO2 levels greater than 60 had a higher chance of survival.

Check whether there is an association between demographic factors(age, gender, & race) and vital.status. Chi-square test will be used to determine whether there is an association between categorical variables. 
```{r}
#Vitals & Age
#Create a table to understand the number of values in each section
vitals.age = table(icu$vital.status, icu$age,
                     dnn = c("Vital Status", "Age"))
addmargins(vitals.age)
chisq.test(vitals.age)
```
Since the p-value(0.005165) from the chi-square test is less than 0.05, it allows us to reject our null hypothesis and we can determine that age does have an association with vital status.

```{r}
#Vitals & Gender
#Create a table to understand the number of values in each section
vitals.gender = table(icu$vital.status, icu$gender,
                     dnn = c("Vital Status", "Gender"))
addmargins(vitals.gender)
chisq.test(vitals.gender)
```
The p-value for vitals and gender is 0.913, which is not less than 0.05 so we cannot reject our null hypothesis. This means that there is no association between gender and vital status. The p-value is extremely high which does not mean significance.  

```{r}
#Vitals & Race
#Create a table to understand the number of values in each section
vitals.race = table(icu$vital.status, icu$race,
                     dnn = c("Vital Status", "Race"))
addmargins(vitals.race)
chisq.test(vitals.race)
```
The p-value for vitals and gender is 0.4046, which is not less than 0.05 so we cannot reject our null hypothesis. This means that there is no association between race and vital status.

P-values:
  Age -> 0.005165
  Gender -> 0.913
  Race -> 0.4046  

It can be concluded that age is the only demographic factor that impacts the vital status of an ICU patient. This makes logical sense because as patients are older they are more susceptible to other underlying conditions that can be fatal.    

The Arterial Blood Gas(ABG) values (pH, PO2, PCO2, & bicarbonate) will be checked in order to see whether there is an association with vital.status. 
```{r}
#ABG & Vital 
#Vitals & PO2
vitals.po2 = table(icu$vital.status, icu$PO2,
                     dnn = c("Vital Status", "Oxygen Partial Pressure"))
addmargins(vitals.po2)
chisq.test(vitals.po2)
```

The p-value for vitals and PO2 is 0.3969, which is not less than 0.05 so we cannot reject our null hypothesis. This means that there is no association between PO2 and vital status.

```{r}
#Vitals & PH
vitals.ph = table(icu$vital.status, icu$PH,
                     dnn = c("Vital Status", "Normal Blood pH"))
addmargins(vitals.ph)
chisq.test(vitals.ph)
```

The p-value for vitals and pH is 0.5187, which is not less than 0.05 so we cannot reject our null hypothesis. This means that there is an association between pH and vital status.

```{r}
#Vitals & PCO2
vitals.pco2 = table(icu$vital.status, icu$PCO2,
                     dnn = c("Vital Status", "Carbon Dioxide Partial Pressure"))
addmargins(vitals.pco2)
chisq.test(vitals.pco2)
```
The p-value for vitals and PCO2 is 1, which is not less than 0.05 so we cannot reject our null hypothesis. This means that there is no association between PCO2 and vital status.

```{r}
#Vitals & bicarb
vitals.bicarb = table(icu$vital.status, icu$bicarb,
                     dnn = c("Vital Status", "Bicarbonate Level"))
addmargins(vitals.bicarb)
chisq.test(vitals.bicarb)
```
The p-value for vitals and bicarb is 0.3141, which is not less than 0.05 so we cannot reject our null hypothesis. This means that there is no association between bicarbonate levels and vital status.

P-values:
  PO2 -> 0.3969
  PH -> 0.5187
  PCO2 -> 1
  bicarb -> 0.3141
  
Based off of the chi-square tests, none of the ABG measures are associated with vital status because all of the p-values are greater than 0.05. The conclusion can be made that a patients outcome after visiting the ICU will not be significantly affected by PO2, PH, PCO2, or bicarbonate levels. 

The other general factors that can impact vital status will also be considered and their association with the predictor variable will be analyzed. 
```{r}
#Vitals & Conscious
vitals.cons = table(icu$vital.status, icu$conscious,
                     dnn = c("Vital Status", "Conscious"))
addmargins(vitals.cons)
chisq.test(vitals.cons)
```
The p-value for vitals and conscious is 1.091e-10, which is much smaller than 0.05 so we can reject our null hypothesis. This means that there is an association between conscious level and vital status.
```{r}
#Vitals & Previous
vitals.prev = table(icu$vital.status, icu$previous,
                     dnn = c("Vital Status", "Previous"))
addmargins(vitals.prev)
chisq.test(vitals.prev)
```
The p-value for vitals and previous is 0.8045, which is not less than 0.05 so we cannot reject our null hypothesis. This means that there is no association between previous and vital status.

```{r}
#Vitals & Service
vitals.serv = table(icu$vital.status, icu$service,
                     dnn = c("Vital Status", "Service"))
addmargins(vitals.serv)
chisq.test(vitals.serv)
```
The p-value for vitals and service is 0.01446, which is less than 0.05 so we can reject our null hypothesis. This means that there is an association between service and vital status.
```{r}
#Vitals & Type
vitals.typ = table(icu$vital.status, icu$type,
                     dnn = c("Vital Status", "Type"))
addmargins(vitals.typ)
chisq.test(vitals.typ)
```
The p-value for vitals and type is 0.001177, which is less than 0.05 so we can reject our null hypothesis. This means that there is an association between type and vital status.

```{r}
#ICU & Chronic Renal Failure
icu.renal = table(icu$vital.status, icu$renal,
                     dnn = c("Vital Status", "Renal"))
addmargins(icu.renal)
chisq.test(icu.renal)
```
The p-value for vitals and renal is 0.0257, which is less than 0.05 so we can reject our null hypothesis. This means that there is an association between renal and vital status.
```{r}
#ICU & Probable Infection
icu.infect = table(icu$vital.status, icu$infect.prob,
                     dnn = c("Vital Status", "Probable Infection"))
addmargins(icu.infect)
chisq.test(icu.infect)
```
The p-value for vitals and infection is 0.01641, which is less than 0.05 so we can reject our null hypothesis. This means that there is an association between renal and vital status.


Since vital.status is a categorical variable, multinomial logistic regression will be used to understand the relationship between the predictor variables and the response variable. The four predictor variables have already been checked to be significant using the chi-square test.  
```{r}
summary(glm(vital.status ~ age + service + renal + infect.prob, data = icu, family=binomial(link = "logit")))
```
Regression Equation:
log_odds = -2.96348 + 0.02559*age + -0.80716*servicesurgical + 0.80719*renalyes + 0.60780*infect.probyes

The p-values for two variable in this model is statistically significant, so a better model is required.  


```{r}
#Logistic regression will be used because vital status is a binary variable
summary(glm(vital.status ~ renal + infect.prob + age, data = icu, family = binomial(link = "logit")))
```
Regression Equation:

log_odds = -3.29684 + 0.94835*renalyes + 0.77453*infect.probyes + 0.02329*age

It is important to note that renalyes has a p-value that is slightly above 0.05, so it is not completely significant. 

```{r}
#The only value that needs to be changed is infection probable, so the other values are arbitrary as long as both patients are the same otherwise
#Each variable will have a 0 if a patient has that attribute and 0 if they do not

#Odds of Mortality given Infection Probable
log_odds_inf = -3.29684 + 0.94835*0 + 0.77453*1 + 0.02329*20
odds_inf = exp(log_odds_inf)
paste("log odds: ", log_odds_inf, sep=" ")
paste("odds: ", odds_inf, sep=" ")

#Odds of Mortality given No Infection Probable
log_odds_no = -3.29684 + 0.94835*0 + 0.77453*0 + 0.02329*20
odds_no = exp(log_odds_no)
paste("log odds: ", log_odds_no, sep=" ")
paste("odds: ", odds_no, sep=" ")
```
We can determine that if a patient is 20 years of age and has a probable infection, they will have 0.12789 odds of living, whereas if that patient does not have a probable infection their odds of living is only 0.05895. The odds ratio is greater than 1 which means that the higher risk is associated with the group of people that have the infection. This does makes sense because someone that is at a risk of having an infection might experience complications that could lead to death. 
